### 2019-10-22 Manuscript/manuscript.tex
regenerate figures and make new reconstruction figure prior to submission

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age. For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below).

@@

\subsection{Discussion}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}.}}}
\label{fig:Laurentia_reconstructions}
\end{figure} 
%ADD MORE RECONSTRUCTIONS. 1 per 100 million years perhaps with indication of which ones are well-constrained by models and

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

@@

The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles. As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian.

There is strong evidence both in the geological and paleomagnetic record of Laurentia for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. As an example, consider the record of Laurentia in the late Mesoproterozoic. As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma. These poles

>>>>

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age. For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2). Additional not-rated poles included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Chuar Group as presented in \cite{Eyster2019a} are also included.

@@

\subsection{Comparing paleogeographic models to the paleomagnetic compilation}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

@@

\subsection{Conclusion}

There is strong evidence both in the geological and paleomagnetic record of Laurentia for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. As an example, consider the record of Laurentia in the late Mesoproterozoic. As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma. These poles constrain rapid motion of Laurentia along

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}. The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

The paleogeographic record of Laurentia is rich in constraints through the Precambrian both in terms of the geological and geochronological constraints on tectonism and the record of paleomagnetic poles. As can be seen in the Chapters on Archean paleogeography (Salminen et al., this volume), Nuna (Elming et al., this volume) and Rodinia (Evans et al., this volume), these constraints are at the center of developing paleogeographic models through the Precambrian and will continue to be moving forward.

### 2019-10-22 Manuscript/pole_table.tex
regenerate figures and make new reconstruction figure prior to submission

            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     279.0 &      56.0 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ... \\

@@

                     Laurentia &                        Ontario lamprophyre dykes &      A &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\

@@

@@

                     Laurentia &                                  Tsezotene Sills &      B &     235.0 &      63.5 & 137.8 &   1.6 &       5.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                                   \textbackslash cite\{Park1989a\} \\
                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 129.4 &  13.9 &       8.2 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Nordic workshop calculation based on data of \textbackslash ... \\
                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &    775\$\textasciicircum \{+25\}\$\$\_\{-25\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    760\$\textasciicircum \{+30\}\$\$\_\{-30\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      724\$\textasciicircum \{+3\}\$\$\_\{-3\}\$ &                               \textbackslash cite\{Denyszyn2009a\} \\

>>>>

            Laurentia-Superior &                       Haig/Flaherty/Sutton Mean  &      B &     281.0 &      56.2 & 245.8 &   1.0 &       3.9 &     1870\$\textasciicircum \{+1\}\$\$\_\{-1\}\$ &  Nordic workshop calculation based on data of \textbackslash ... \\

@@

                     Laurentia &                        Ontario lamprophyre dykes &     NR &     273.3 &      48.8 & 223.3 &  58.0 &       9.2 &   1143\$\textasciicircum \{+10\}\$\$\_\{-10\}\$ &                                 \textbackslash cite\{Piispa2018a\} \\

@@

                     Laurentia &                      Adirondack fayalite granite &     NR &     285.5 &      44.0 & 132.7 & -28.4 &       6.9 &    990\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
                     Laurentia &             Adirondack metamorphic\textbackslash nanorthosites &     NR &     286.0 &      44.0 & 149.0 & -25.1 &      11.6 &    970\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\
                     Laurentia &                     Adirondack Microcline gneiss &     NR &     285.0 &      44.0 & 151.1 & -18.4 &      10.5 &    960\$\textasciicircum \{+20\}\$\$\_\{-20\}\$ &                                  \textbackslash cite\{Brown2012a\} \\

@@

                     Laurentia &                                  Gunbarrel dykes &      B &     248.7 &      44.8 & 138.2 &   9.1 &      12.0 &      778\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &  Calculation from \textbackslash cite\{Eyster2019a\} based on d... \\
            Laurentia-Svalbard &                        Svanbergfjellet Formation &      B &      18.0 &      78.5 & 226.8 &  25.9 &       5.8 &    770\$\textasciicircum \{+19\}\$\$\_\{-40\}\$ &                                 \textbackslash cite\{Maloof2006a\} \\
                     Laurentia &                             Uinta Mountain Group &      B &     250.7 &      40.8 & 161.3 &   0.8 &       4.7 &     760\$\textasciicircum \{+6\}\$\$\_\{-10\}\$ &                                   \textbackslash cite\{Weil2006b\} \\
                     Laurentia &                                    Carbon Canyon &     NR &     248.2 &      36.1 & 166.0 &  -0.5 &       9.7 &      757\$\textasciicircum \{+7\}\$\$\_\{-7\}\$ &  \textbackslash cite\{Weil2004a\} as calculated in \textbackslash cite\{Eyster... \\
                     Laurentia &                             Carbon Butte/Awatubi &     NR &     248.5 &      35.2 & 163.8 &  14.2 &       3.5 &      751\$\textasciicircum \{+8\}\$\$\_\{-8\}\$ &                                 \textbackslash cite\{Eyster2019a\} \\
                     Laurentia &                        Franklin event grand mean &      A &     275.4 &      73.0 & 162.1 &   6.7 &       3.0 &      718\$\textasciicircum \{+2\}\$\$\_\{-2\}\$ &                               \textbackslash cite\{Denyszyn2009b\} \\
